### 2022-10-10 Proposal./proposal.tex
Uploading proposal

\documentclass[12pt]{article}
\usepackage{filecontents}
\usepackage{natbib}

\title{Proposal}

% Add author information below. Communicating author is indicated by an asterisk, the affiliation is shown by superscripted lower case letter if several affiliations need to be noted.
\author{
Owen Fiore
}

\pagestyle{empty}

\begin{document}

\section{Introduction}

At the 2022 World Track Championships in Eugene Oregon, Devon Allen was disqualified in the
Men’s 110 meter hurdle final after registering a reaction time of 0.099 seconds, 0.001 
seconds faster than what is allowed.  Following the games, bloggers on the running website 
LetsRun scrutinized the results and performed non-statistical analysis of the data \citep{Johnson}
They found that the reaction time data from the 2022 World Track Championships seemed to be faster 
compared to the other datasets they looked at, but did not perform any meaningful statistical 
analysis of the data.  Devon Allen had previously ran the third fastest time in history when 
he ran 12.84 earlier in 2022 and was considered to be one of the favorites to win \citep{Preview}.  His 
disqualification thus prompted questions into whether the timing equipment used at the World
Track Championships was accurate and whether or not Allen actually reacted as fast as the 
timing equipment stated he did.

\section{Specific Aims}
I plan to investigate whether the timing equipment used at the Track and Field World Championships
 produced reaction times different than what times typically are.  Since 1985, Seiko has timed 
 every World Championships with 2022 being no exception\citep{Seiko}.  Thus it would be surprising to discover
 if the timing equipment was producing faster than actual times as the same company was responsible
 for timing the Championships.  By comparing the timing data from 2022 to other World Championship 
 meets, it will be established whether 2022 was an outlier and whether Devon Allen was wrongly 
 disqualified.

\section{Data}

The data set I will be looking at comes from the World Athletics Website which has recorded data
for reaction times in the Men’s 110 meter hurdle since 1999. There are over 800 observations for
reaction times in the heats, semi-finals, and finals for the hurdles since 1999 and there were
67 in 2022.  I intend to look at variables such as year, stage of competition (heats vs finals),
reaction time, total time, and if possible attempt to find environmental data looking at 
temperature, humidity, and elevation.  The data can be found here: \citet{WAData}

\section{Methods}
I plan to use Tukey’s Honestly Significant Difference test to compare pairs of data by year
going back to 1999 to see if any other World Track Championship had significantly different 
results.  Using linear prediction regression in R, I intend to look at whether for each 
championship whether the reaction times got faster in each successive stage of competition.
By comparing data from the heats to the semifinals to the finals, I hope to get a better 
understanding of any confounding variables.  There is also data available for the USA Track
Championships which occurred in June of 2022.  By comparing runners who ran in both the USATF
Championships and the World Championships, a matched pairs t test can be conducted.  Lastly, I
will use a linear regression model that looks at reaction time and the following independent
variables: Year, final time, stage of competition, and any data I can find on the weather at the
meet.

\section{Discussion}
I hope to find data that shows that the 2022 World Championships reaction times were significantly
lower than any other World Championship and that the most significant predictor in determining
reaction time is the year the data was taken.  I have already looked at a very brief sample set
comparing the data from 2019 to 2022 and found that the 2022 reaction times were statistically
significantly faster than in 2019.  If my investigation results in evidence that suggests that
Devon Allen was disqualified due to faulty data, and my work is published I hope it changes the
IAAF ruling of immediately disqualifying all of those who false start, especially under such a
small margin.  If the data is not what I expect, then I do not anticipate much change to occur as
a result.

\bibliographystyle{chicago}
\bibliography{citations.bib}

\end{document}

### 2022-11-14 manuscript/manuscript.tex
First draft

\documentclass[12pt, letterpaper, titlepage]{article}

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{colorlinks = true, linkcolor = blue, citecolor=blue, urlcolor = blue}
\usepackage{natbib}

\usepackage{setspace}

\usepackage[pagewise]{lineno}
%\linenumbers*[1]
% %% patches to make lineno work better with amsmath
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
 {\linenomath\csname old#1\endcsname}%
 {\csname oldend#1\endcsname\endlinenomath}}%
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%

\AtBeginDocument{%
 \patchBothAmsMathEnvironmentsForLineno{equation}%
 \patchBothAmsMathEnvironmentsForLineno{align}%
 \patchBothAmsMathEnvironmentsForLineno{flalign}%
 \patchBothAmsMathEnvironmentsForLineno{alignat}%
 \patchBothAmsMathEnvironmentsForLineno{gather}%
 \patchBothAmsMathEnvironmentsForLineno{multline}%
}

% control floats
\renewcommand\floatpagefraction{.9}
\renewcommand\topfraction{.9}
\renewcommand\bottomfraction{.9}
\renewcommand\textfraction{.1}
\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}

\newcommand{\jy}[1]{\textcolor{blue}{JY: #1}}
\newcommand{\eds}[1]{\textcolor{red}{EDS: (#1)}}

\title{Was Devon Allen Unjustly Disqualified at the 2022 World Track and Field Championships?}

\author{Owen Fiore\\
%   \href{mailto:owen.fiore@uconn.edu}
% {\nolinkurl{owen.fiore@uconn.edu}}\\
  Elizabeth Schifano\\
  Jun Yan\\[1ex]
  Department of Statistics, University of Connecticut\\
}
\date{}

\begin{document}
\maketitle

\doublespace

\begin{abstract}
  At the 2022 World Track Championships in Eugene Oregon, Devon Allen was disqualified in the
  Men's 110 meter hurdle final after registering a reaction time of 0.099 seconds, 0.001 
  seconds faster than what is allowed.  Following the games, bloggers on the running website 
  LetsRun scrutinized the results and performed non-statistical analysis of the data \citep{Johnson}
  They found that the reaction time data from the 2022 World Track Championships seemed to be faster 
  compared to the other datasets they looked at, but did not perform any meaningful statistical 
  analysis of the data.  Devon Allen had previously ran the third fastest time in history when 
  he ran 12.84 earlier in 2022 and was considered to be one of the favorites to win \citep{Preview}.  His 
  disqualification thus prompted questions into whether the timing equipment used at the World
  Track Championships was accurate and whether or not Allen actually reacted as fast as the 
  timing equipment stated he did.

\bigskip
\noindent\sc{Keywords}:
Reaction Time, Seiko, 2022 World Track and Field Championships, Devon Allen

\end{abstract}

\section{Introduction}
\label{sec:intro}

The contribution of this paper is to question whether or not Seiko's timing data
was accurate and thus if Devon Allen was disqualified more so because of faulty
equipment rather than reacting illegally fast.  There is little at this point
that could be done to rectify any potential mis-qualification other than to try
to prevent it from happening again.

One of the most interesting things found was that the reaction times from the USA
athletes in the World Championship 110 meter hurdles, was that they were significantly
faster than at the USA Track Championships.  From June 23-26 2022, USA held it's Track
and Field Championship to decide who to send to the World Championships.  Thus, a baseline
is able to be established for the four USA athletes who competed in at least one round
of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen. Every
athlete reacted faster in all of the World Track and Field Championship races compared
to the USA Track and Field Championship races. Devon Allen raced three times at
the USA Track and Field Championship and three times at the World Track and Field 
Championship.  His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teamates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  This table below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. "USA" denotes
the USA Track and Field Championships, "World" denotes the World Track and Field
Championships, "H" denotes the heats/preliminary rounds of the 110 meter hurdles,
"S" denotes the semifinals of the 110 meter hurdles, "F" denotes the finals of the
110 meter hurdles.  "N/A" was used for any athlete that did not compete and thus
did not register a reaction time.  All numbers listed in the table are the reaction
time for the respective athletes in seconds.

\begin{figure}
    \begin{center}
      \begin{tabular}{||c | c c c | c c c||} 
       \hline
       Athlete & USA H & USA S & USA F & World H & World S & World F \\ [0.5ex] 
       \hline\hline
       Devon Allen & 0.201 & 0.153 & 0.160 & 0.123 & 0.101 & 0.099 \\ 
       \hline
       Trey Cunningham & 0.186 & 0.185 & 0.182 & 0.115 & 0.120 & 0.109 \\
       \hline
       Grant Holloway & 0.192 & 0.190 & N/A & 0.147 & 0.128 & 0.124 \\
       \hline
       Daniel Roberts & 0.181 & 0.200 & 0.183 & 0.179 & N/A & N/A \\ [0.5ex]
       \hline
      \end{tabular}
      \end{center}

\end{figure}

Additionally it is worth noting that in the semi-finals of the World Track and
Field Championships; Allen's reaction time was 0.101 which is only 0.001 above
the legal limit.  What this suggests is that Allen may have strong reflexes
and be able to react extremely well to the sound of the gun.  Devon Allen is a
very good athlete, as evidenced by his football career at the University of Oregon
and making it onto the Philadelphia Eagles practice squad this past year \citep{Hurley}.
Thus the 0.099 reaction time may not have been a product of Devon Allen
predicting the start of the race but rather a combination of a quick reaction 
and a possibly faulty sensor.

Since 1985 Seiko Holding Corporations has served every World Athletics Championship
as the official timer \citep{Seiko}.  Since 1985, the technology and the ability
to accurately predict measurements: long jump distances, false starts, total time,
reaction time, etc. have all dramatically improved.  Seiko did not start tracking
reaction time as an official measurement until 1999 in the men's and women's 100 meter
hurdles.  Seiko regularly updates their equipment so that they provide cutting edge
technology to the World Track and Field Championships, the highest stakes in the world
of running outside of the Olympics.

%Citation needed in this section
Seiko's technology for detecting reaction times relies on the pressure that athletes
exert on the plate when they push off.  Their systems measure the time differential
between when the "gun" goes off to start the race and when the pressure changes 
\citep{Seiko}  If an athlete has a reaction time under 0.10 seconds, it is deemed a 
false start as it is considered that no athlete can react so quickly \citep{Seiko-Timing}.  
Thus, a reaction time of for example 0.05, suggests that the athlete predicted the gun and it
was luck that caused their abnormally high time.  At the World Championship level,
World Athletics and Seiko want to remove that element of luck and thus impose the 0.1
second barrier.

In 2013, Seiko upgraded their timing equipment for sprinting events (100 meter hurdles
falls under this catergory) \citep{WorldAthletics_2013}.  Since 1999, the highest reaction 
times in the men's 100 meter hurdle were recorded in 2013.  That is not to say that the higher 
times were caused by Seiko's equipment, but the two may be related.  It is worth noting that prior
to 2022, Seiko again upgraded its technology but for its jump management system.
\citep{Pilianidis}

\section{Data}
\label{sec:data}
Data was copied from the World Athletics website and pasted into an excel
spreadsheet. The data covers the men's 110 meter hurdles and the women's 100
meter hurdles from 1999 to 2022.  The variable of prediction interest is reaction
time, and the predicting variables are year, stage of competition (heats, 
semifinal, final), total time, and gender.  The data can be found in the appendix:.  
Factors such as name, country, and identification number were
not considered important for the purposes of this research besides the times for
Devon Allen and the other United States athletes mentioned above.  As the reaction
times for men and women did not differ significantly, the results were pooled
to allow for larger sample sizes, as the data from the finals of each year is
typically very small (around 8 observations), so including men and women doubles
the sample size per year.  This is also very useful because t tests are sometimes
inconsistent and unreliable when the sample size is small.

\section{Methods}
\label{sec:methods}
The main purpose of this paper is to determine whether or not Devon Allen was
unfairly disqualified because of his reaction time. By comparing the results from
2022 to other years, we can hope to learn more about what causes athletes to react
the way in which they do.  A very simple way to compare reaction time across the many
years was to perform a Tukey's Honestly Significant Test to test all pairs of
years against one another to see what years had reaction time data that differed
significantly from another year.  Tukey's HSD performs a series of t tests, with
each t test for one pair of years, with the reaction time being the variable of
interest.  Tukey's HSD relies on the following assumptions: observations are
idendependent amongst groups, groups are normally distributed, each group has
homogenity of variance \citep{statistics}.
Another important and slightly more complex
method is to use a linear mixed effects model.  A LME model is used when there
are groups that can cause an effect on the dependent variable when not properly
detected.  For example, a scatterplot may show that increasing execerise
increases the chance of a heart attack, but what is not shown is that age is a
factor that affects how much people exericse and if older people exercise more
and are more at risk for a heart attack, then it would incorrect to conclude
that exercise is what is causing the heart attacks.  In this case, it would be said
that there is an "age factor" associated with heart attacks that needs to be
accounted for.
In the paper, we need to determine if there is a year effect that can impact
the reaction time.  In R, this can be accomplished with the lmer() function
in the lme4 package.  Then, the LME model can be compared to see if it differs
significantly from a model that just looks at whether there is a random effect
of year on reaction time.  This can be done using the anova() function, which
is used to compare different models.  We expect that there is a year effect,
meaning that the anova test would produce a p value that is statistically
significant.  It would not make sense to fit a traditional linear regression
model to try and predict the reaction time based off the year, instead we want
to stratify by year and look at the reaction time to see if there is a year
effect in determining reaction time.
The data used will primarily be the finals data because that is most relevant
to Devon Allen's disqulification and because of differences in times compared
to the semi-finals and heats.  In the finals, runners may be more nervous and
have more adrenlaine and may be more "jumpy" which would cause them to be
able to better react to the start of the gun, rather than in the preliminary
rounds.

\section{Conclusion}
\label{sec:conclusion}
The Tukey HSD performed on the finals data showed that across the many t tests it
performed, 8 were signifcantly different of which 6 came from 2022 with the other
two came from 2013 which was a high outlier on reaction time.  No other year had
as low as a reaction time as 2022 for the finals, and the individual t test
comparing 2019 and 2022 produced a p value of $2.654\cdot10^{-8}$.  This is an
extermely significant p value and shows that there was a significant difference
betweent the 2019 times and 2022 times, but there does not appear to be a reason
for why the times should be so much higher.  Shoe technology may have improved
during that time, but that would likely be reflected in the total time and not
in the reaction time.  Additionally, other environmental factors such as wind,
heat, etc. also should not impact how quick athletes are to react to the sound
of the gun.  When the analysis of variance was run to compare the linear mixed
effects model and the random effects model, the resulting p value was $3.243\cdot10^{-5}$.
This is a highly significant p value at all alpha levels and it means that there
does appear to be a year effect when predicting reaction time.  From the R output,
the estimate of the intercept was 0.148883, the variance of the intercept was
$8.643\cdot10^{-5}$, the variance of the residuals was $3.719\cdot10^{-4}$.
However, this does not provide sufficient evidence of how strong the year effect
is.  In order to calculate that, the pnorm() function was used, which returned
a p value of 0.0112, which is low but not significant at all alpha levels.  This
means that there is a very great chance that the year effect is significant but
that we expect that if we repeatly sampled, the chance of observing a test
statistic higher would be only 0.0112.

Devon Allen's disqulification at the World Track and Field Championships was
possibly the result of both a faulty timing equipment and Devon Allen
reacting extremely quickly to the start.  The code and data from R showed that
the 2022 World Track and Field Championships were a low outlier compared to
every other year in terms of average reaction time.  Addtionally, it was shown
through a linear mixed effects model that there is a year effect that is significant
in determining reaction time.  However, this practically does not make much sense
as there is no reason for so much random variation without much of a trend.
There has not been a consistent decrease or increase since 1999, much of the data
for reaction times has been random and unpredictable.  Thus while it seems easy
to conclude that Devon Allen was wrongly disqualified, that may not neccessarily
be the case.

\section{Appendix}
\label{sec:appendix}
%Put downloadable data spreadsheet
Here is World Athletics website with the data: \url{https://www.worldathletics.org/results/world-athletics-championships}.
Here is the data for the USA Track and Field Championships: \url{https://www.flashresults.com/2022_Meets/Outdoor/06-23_USATF/}

\bibliographystyle{chicago}
\bibliography{citations.bib}

\end{document}

### 2022-11-14 manuscript/manuscript.tex
Minor changes to intro section

@@
 that could be done to rectify any potential mis-qualification other than to try
-to prevent it from happening again.
+to prevent it from happening again. Devon Allen's disqulification needs to be
+addressed because this is a problem that could continue to re-appear.  If athletes
+are reacting fast enough that the timing system thinks that they are reacting
+illegally fast, then World Athletics needs to address this and possibly change
+the minimum reaction time allowed.  Other papers have discussed track and field
+reaction times but not in the context of whether the minimum reaction time barrier
+is too low or whether or not timing systems were at fault.
 

### 2022-11-14 Code/ReactionTime.Rmd
Added code for project

---
  title: "Thesis"
author: "Owen Fiore"
date: "10/20/2022"
output: pdf_document
---

Reading in the Data
```{r}
library(readxl)
times <- read_excel("C:/Users/ofior/Documents/Thesis/Data/ThesisData.xlsx")
View(times)
colnames(times) <- c("Year", "Type", "Time", "RxnTime", "Gender")
```

t test comparing 2019 and 2022
```{r}
t.test(times$RxnTime[times$Year==2019],times$RxnTime[times$Year==2022])
```

Creates a Boxplot comparing 2019 and 2022
```{r}
boxplot(times$RxnTime[times$Year==2019],times$RxnTime[times$Year==2022])
```

Analysis of Variance, looking only at Year
```{r}
model <- aov(RxnTime ~ Year, data = times)
summary(model)
```

Creates a new data set of just finals and looks at the data from that
```{r}
finals <- subset(times, Type == "F")
boxplot(RxnTime ~ Year, data = finals, xlab = "Years")

```

```{r}
library(dplyr)
finals.aov <- aov(RxnTime ~ as.factor(Year), data = finals)
summary(finals.aov)
tuk.final <- TukeyHSD(finals.aov, conf.level = .95)
tkf <- as.data.frame(tuk.final[1])
tkf %>%filter(as.factor.Year..p.adj < 0.05)
#plot(tuk.final)
```

```{r}
residuals <- resid(finals.aov)
plot(finals$RxnTime, residuals, xlab="RxnTime", ylab="Residuals")
abline(0,0)
```

```{r}
library(lme4)
finals.lme <- lmer(RxnTime ~ (1|Year), data = finals)
summary(finals.lme)
qqnorm(resid(finals.lme))
hist(resid(finals.lme))
qqnorm(ranef(finals.lme))
pnorm(0.1, mean = .148883, sd = sqrt(8.648*10^(-5)+3.719*10^(-4)))
```
```{r}
library(lme4)
semi <- subset(times, Type == "S")
finals.lme <- lmer(RxnTime ~ (1|Year), data = semi)
summary(finals.lme)
qqnorm(resid(finals.lme))
hist(resid(finals.lme))
qqnorm(ranef(finals.lme))
```
```{r}
library(lme4)
heat <- subset(times, Type == "H")
finals.lme <- lmer(RxnTime ~ (1|Year), data = heat)
summary(finals.lme)
qqnorm(resid(finals.lme))
hist(resid(finals.lme))
qqnorm(ranef(finals.lme))
str(ranef(finals.lme))
hist(ranef(finals.lme)$Year[,1])
qqnorm(ranef(finals.lme)$Year[,1])
qqline(ranef(finals.lme)$Year[,1])
```
```{r}
pnorm(0.1, mean = .157750, sd = sqrt(.0001228+0.0005937))
qnorm(0.01, mean = .157750, sd = sqrt(.0001228+0.0005937))
```

```{r}
my_fit <- lm(RxnTime ~ 1 , data = finals)
summary(my_fit)
```

```{r}
anova(finals.lme, my_fit)
```

```{r}
old <- subset(times, Year < 2020)
old.lme <- lmer(RxnTime ~ (1|Year), data = old)
summary(old.lme)
old_fit <- lm(RxnTime ~ 1 , data = old)
anova(old.lme, old_fit)

```

```{r}
model_coefs <- coef(old.lme)$Year

model_coefs
```

Plotting LME
```{r}
library(ggplot2)
model_coef_plot <- ggplot(data = old, 
                          mapping = aes(x = Time, 
                                        y = RxnTime, 
                                        colour = Year)
) +
  scale_y_continuous(limits = c(180, 1020)) +
  scale_x_continuous(breaks = seq(1:10) - 1) +
  theme(legend.position = "top")

model_coef_plot
```

Creates a new data set of just semifinals and looks at the data from that
```{r}
semi <- subset(times, Type == "S")

semi.fit <- lm(RxnTime ~ Year, data = semi)
summary(semi.fit)
```

```{r}

boxplot(RxnTime ~ Year, data = semi, xlab = "Years")
```

```{r}

semi.aov <- aov(RxnTime ~ as.factor(Year), data = semi)
summary(semi.aov)
tuk.semi <- TukeyHSD(semi.aov, conf.level = .95)
tks <- as.data.frame(tuk.semi[1])
tks %>%filter(as.factor.Year..p.adj < 0.05)

```
```{r}
residuals.semi <- resid(semi.aov)
plot(semi$RxnTime, residuals.semi, xlab="RxnTime", ylab="Residuals")
abline(0,0)
## Off by 1 error
```

Creates a new data set of just heats and looks at the data from that
```{r}
heat <- subset(times, Type == "H")

heat.fit <- lm(RxnTime ~ Year, data = heat)
summary(heat.fit)
```

```{r}

boxplot(RxnTime ~ Year, data = heat, xlab = "Years")
```

```{r}

heat.aov <- aov(RxnTime ~ as.factor(Year), data = heat)
summary(heat.aov)
tuk.heat <- TukeyHSD(heat.aov, conf.level = .95)
tkh <- as.data.frame(tuk.heat[1])
tkh %>%filter(as.factor.Year..p.adj < 0.05)

```

t test comparing men and women times in 2022
```{r}
men <- subset(times, Gender == "M")
women <- subset(times, Gender == "F")
t.test(men$RxnTime, women$RxnTime)

```

Scatterplot of Data
```{r}
plot(finals$Year, finals$RxnTime)
```
```{r}
ninetynine <- times$RxnTime[times$Year == "1999"]
one <- times$RxnTime[times$Year == "2001"]
three <- times$RxnTime[times$Year == "2003"]
five <- times$RxnTime[times$Year == "2005"]
seven <- times$RxnTime[times$Year == "2007"]
five <- times$RxnTime[times$Year == "2005"]
thirteen <- times$RxnTime[times$Year == "2013"]
two <- times$RxnTime[times$Year == "2022"]
```

```{r}
plot(times$RxnTime, times$Time)
plot(times$RxnTime[times$Time < 15], times$Time[times$Time < 15], xlab = "RxnTime", ylab = "Total Time")
```

### 2022-12-09 manuscript/manuscript.tex
Fixed table
significant revisions

--- 2022-11-14 manuscript/manuscript.tex
@@
 \usepackage{natbib}
-
+\usepackage{float}
 \usepackage{setspace}
@@
 \noindent\sc{Keywords}:
-Reaction Time, Seiko, 2022 World Track and Field Championships, Devon Allen
+Devon Allen, Reaction Time, Seiko, 2022 World Track and Field Championships
 
@@
 
+Other scholarly papers have discussed track and field reaction times but not in the
+context of whether the minimum reaction time barrier is too low or whether or not timing
+systems were accurate.  Outside of the blog posts from LetsRun.com, there have not been
+any published papers regarding Devon Allen's disqualification in particular.  This is not
+neccessarily suprising given that the World Track and Field Championships occurred only four months ago.
+
 The contribution of this paper is to question whether or not Seiko's timing data
@@
 that could be done to rectify any potential mis-qualification other than to try
-to prevent it from happening again.
+to prevent it from happening again.  THis matter needs to be addressed because Devon Allen's disqulification could be a problem that could continue to re-appear.  If athletes
+are reacting fast enough that the timing system thinks that they are reacting
+illegally fast, then World Athletics needs to address this and possibly change
+the minimum reaction time allowed.  Currently the barrier is 0.10 seconds, any time
+faster than this would result in a disqualification in the athlete who reacted too
+quickly and the race is restarted \citep{Seiko-Timing}.
+Any athletes who false start are not given a second chance, they are removed from competition.
+This can have massive implications as events like the Outdoor World Track and Field Championship only occur typically once every two years.
+It is an opportunity for athletes to earn endorsement deals and prove to sponsors such as Nike and Adidas that
+they deserve a contract.  Thus the significance of a disqualification, especially in the finals of a major event such as the World Track and Field Championships, are very
+consequential.
 
-One of the most interesting things found was that the reaction times from the USA
-athletes in the World Championship 110 meter hurdles, was that they were significantly
-faster than at the USA Track Championships.  From June 23-26 2022, USA held it's Track
-and Field Championship to decide who to send to the World Championships.  Thus, a baseline
-is able to be established for the four USA athletes who competed in at least one round
-of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen. Every
-athlete reacted faster in all of the World Track and Field Championship races compared
-to the USA Track and Field Championship races. Devon Allen raced three times at
-the USA Track and Field Championship and three times at the World Track and Field 
-Championship.  His reaction times were significantly faster in the World Track and
-Field Championship compared to the USA Track and Field Championship as were his
-teamates.  This comparison suggests that the timing system used between the two
-meets may have been meaningfully different to the point where Devon Allen was
-disqualified more because of faulty equipment rather than because he reacted
-too quickly.  This table below highlights the difference in reaction times for
-Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. "USA" denotes
-the USA Track and Field Championships, "World" denotes the World Track and Field
-Championships, "H" denotes the heats/preliminary rounds of the 110 meter hurdles,
-"S" denotes the semifinals of the 110 meter hurdles, "F" denotes the finals of the
-110 meter hurdles.  "N/A" was used for any athlete that did not compete and thus
-did not register a reaction time.  All numbers listed in the table are the reaction
-time for the respective athletes in seconds.
+% \begin{figure}[H]
+%     \begin{center}
+%       \begin{tabular}{||c | c c c | c c c||}[H] 
+%        \hline
+%        Athlete & USA H & USA S & USA F & World H & World S & World F \\ [0.5ex] 
+%        \hline\hline
+%        Devon Allen & 0.201 & 0.153 & 0.160 & 0.123 & 0.101 & 0.099 \\ 
+%        \hline
+%        Trey Cunningham & 0.186 & 0.185 & 0.182 & 0.115 & 0.120 & 0.109 \\
+%        \hline
+%        Grant Holloway & 0.192 & 0.190 & N/A & 0.147 & 0.128 & 0.124 \\
+%        \hline
+%        Daniel Roberts & 0.181 & 0.200 & 0.183 & 0.179 & N/A & N/A \\ [0.5ex]
+%        \hline
+%       \end{tabular}
+%       \end{center}
 
-\begin{figure}
-    \begin{center}
-      \begin{tabular}{||c | c c c | c c c||} 
-       \hline
-       Athlete & USA H & USA S & USA F & World H & World S & World F \\ [0.5ex] 
-       \hline\hline
-       Devon Allen & 0.201 & 0.153 & 0.160 & 0.123 & 0.101 & 0.099 \\ 
-       \hline
-       Trey Cunningham & 0.186 & 0.185 & 0.182 & 0.115 & 0.120 & 0.109 \\
-       \hline
-       Grant Holloway & 0.192 & 0.190 & N/A & 0.147 & 0.128 & 0.124 \\
-       \hline
-       Daniel Roberts & 0.181 & 0.200 & 0.183 & 0.179 & N/A & N/A \\ [0.5ex]
-       \hline
-      \end{tabular}
-      \end{center}
-
-\end{figure}
+% \end{figure}
 
@@
 not considered important for the purposes of this research besides the times for
-Devon Allen and the other United States athletes mentioned above.  As the reaction
-times for men and women did not differ significantly, the results were pooled
-to allow for larger sample sizes, as the data from the finals of each year is
-typically very small (around 8 observations), so including men and women doubles
-the sample size per year.  This is also very useful because t tests are sometimes
-inconsistent and unreliable when the sample size is small.
+Devon Allen and the other United States athletes mentioned above.  The variable "Stage"
+refers to whether the observation occurred during the "heats" (Preliminary round) which
+is denoted by "H" in the data, "S" for "semifinals" (Of which there are either 2 or 3 heats each year), and "F" for finals. The variable "Gender" has levels "M" and "F" which
+stand for "male" and "female" respectively.  The variable "Batch" takes a numerical value and every heat of every round from 1999 to 2022 was assigned it's own number.  This
+was added because of suspected issues that reaction time may be correlated with other runners reacting quickly (one runner reacting quickly may result in other runners reacting quickly).
+
+One of the most interesting things found was that the reaction times from the USA
+athletes in the World Championship 110 meter hurdles, was that they were significantly
+faster than at the USA Track Championships.  From June 23-26 2022, USA held it's Track
+and Field Championship to decide who to send to the World Championships.  Thus, a baseline
+is able to be established for the four USA athletes who competed in at least one round
+of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen. Every
+athlete reacted faster in all of the World Track and Field Championship races compared
+to the USA Track and Field Championship races. Devon Allen raced three times at
+the USA Track and Field Championship and three times at the World Track and Field 
+Championship.  His reaction times were significantly faster in the World Track and
+Field Championship compared to the USA Track and Field Championship as were his
+teamates.  This comparison suggests that the timing system used between the two
+meets may have been meaningfully different to the point where Devon Allen was
+disqualified more because of faulty equipment rather than because he reacted
+too quickly.  This table below highlights the difference in reaction times for
+Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. "USA" denotes
+the USA Track and Field Championships, "World" denotes the World Track and Field
+Championships, "N/A" was used for any athlete that did not compete and thus
+did not register a reaction time.  All numbers listed in the table are the reaction
+time for the respective athletes in seconds.
+
+\begin{center}
+  \begin{tabular}{||c | c c c | c c c||} 
+   \hline
+   Athlete & USA H & USA S & USA F & World H & World S & World F \\ [0.5ex] 
+   \hline\hline
+   Devon Allen & 0.201 & 0.153 & 0.160 & 0.123 & 0.101 & 0.099 \\ 
+   \hline
+   Trey Cunningham & 0.186 & 0.185 & 0.182 & 0.115 & 0.120 & 0.109 \\
+   \hline
+   Grant Holloway & 0.192 & 0.190 & N/A & 0.147 & 0.128 & 0.124 \\
+   \hline
+   Daniel Roberts & 0.181 & 0.200 & 0.183 & 0.179 & N/A & N/A \\ [0.5ex]
+   \hline
+  \end{tabular}
+  \end{center}
 
@@
 The main purpose of this paper is to determine whether or not Devon Allen was
-unfairly disqualified because of his reaction time. By comparing the results from
-2022 to other years, we can hope to learn more about what causes athletes to react
-the way in which they do.  A very simple way to compare reaction time across the many
-years was to perform a Tukey's Honestly Significant Test to test all pairs of
-years against one another to see what years had reaction time data that differed
-significantly from another year.  Tukey's HSD performs a series of t tests, with
-each t test for one pair of years, with the reaction time being the variable of
-interest.  Tukey's HSD relies on the following assumptions: observations are
-idendependent amongst groups, groups are normally distributed, each group has
-homogenity of variance \citep{statistics}.
-Another important and slightly more complex
-method is to use a linear mixed effects model.  A LME model is used when there
-are groups that can cause an effect on the dependent variable when not properly
-detected.  For example, a scatterplot may show that increasing execerise
-increases the chance of a heart attack, but what is not shown is that age is a
-factor that affects how much people exericse and if older people exercise more
-and are more at risk for a heart attack, then it would incorrect to conclude
-that exercise is what is causing the heart attacks.  In this case, it would be said
-that there is an "age factor" associated with heart attacks that needs to be
-accounted for.
-In the paper, we need to determine if there is a year effect that can impact
-the reaction time.  In R, this can be accomplished with the lmer() function
-in the lme4 package.  Then, the LME model can be compared to see if it differs
-significantly from a model that just looks at whether there is a random effect
-of year on reaction time.  This can be done using the anova() function, which
-is used to compare different models.  We expect that there is a year effect,
-meaning that the anova test would produce a p value that is statistically
-significant.  It would not make sense to fit a traditional linear regression
-model to try and predict the reaction time based off the year, instead we want
-to stratify by year and look at the reaction time to see if there is a year
-effect in determining reaction time.
-The data used will primarily be the finals data because that is most relevant
-to Devon Allen's disqulification and because of differences in times compared
-to the semi-finals and heats.  In the finals, runners may be more nervous and
-have more adrenlaine and may be more "jumpy" which would cause them to be
-able to better react to the start of the gun, rather than in the preliminary
-rounds.
+unfairly disqualified because of his reaction time.
+
+The main statistical model used in the paper is a gamma mixed effects model, which is
+in the family of Generalized Linear Mixed Model (GLMM).  A GLMM is a type of model that
+supports inputs that are from different distributions.  The model that we use is for the
+Gamma distribution for the fixed effects part of the model and a random effects part.  In
+the model used in the paper, the fixed effect comes in the form of an intercept, and then
+either a year effect, batch effect, or both are used.  Initially a linear mixed effects
+model was attempted to be implemented, but due to a right hand skew in the reaction times, a gamma distribution seemed better equipped to model the data (The right
+skew is not particularly suprising as it is not possible to register a reaction time below 0.0 seconds, resulting in a lower bound but no upper bound for outliers).
+In the code, the link parameter under Gamma() was changed to be "log" as logarithmic functions are better equipped to deal with high outliers.
+
+There are several different data sets that could be used and explored in determining 
+whether or not Devon Allen was unjustly disqualified.  One of the best examples of this
+is the women's 100 meter hurdles. Although the distance of the race is 10 meters shorter
+than the men's race, the reaction time data is still important and can be used to
+essentially double the data size.  However, one of the largest issues with this is that on average, men are able to react quicker than women.  Thus the data may end up following
+a bimodal distribution which would be difficult to model and not helpful for determining
+if Devon Allen's reaction time was the result of a faulty system or if it is an argument
+for chaning the 0.10 second minimum barrier that currently exists.  Thus it was decided
+to look at men's and women's data separately.  Another decision that was made was to look
+at the data from the semifinals and finals together but from the heats or preliminary rounds.  The reason for this was because it was reasoned that in the finals athletes may
+be a little more anxious, more "jumpy" and that under stress and they would react faster.
+While the preliminary rounds also provide a bulk of the total observations, they are not
+as relevant to Allen's disqualification, which occured during the finals.
+%%%Include picture
 
@@
 \label{sec:conclusion}
+%%%Need to discuss how the false start changed from 2007-2009
 The Tukey HSD performed on the finals data showed that across the many t tests it

### 2022-12-10 manuscript/manuscript.tex
Significant changes

@@
 \usepackage{setspace}
-
+\usepackage{pdfpages}
 \usepackage[pagewise]{lineno}
+\usepackage{mwe}
 %\linenumbers*[1]
@@
 
-\bigskip
+%\bigskip
 \noindent\sc{Keywords}:
@@
 faster than this would result in a disqualification in the athlete who reacted too
-quickly and the race is restarted \citep{Seiko-Timing}.
-Any athletes who false start are not given a second chance, they are removed from competition.
+quickly and the race is restarted \citep{Seiko-Timing}.  Since 2011, any athletes who false start are not given a second chance, they are removed from competition
+\citep{False-Start}.
 This can have massive implications as events like the Outdoor World Track and Field Championship only occur typically once every two years.
@@
 
-% \begin{figure}[H]
-%     \begin{center}
-%       \begin{tabular}{||c | c c c | c c c||}[H] 
-%        \hline
-%        Athlete & USA H & USA S & USA F & World H & World S & World F \\ [0.5ex] 
-%        \hline\hline
-%        Devon Allen & 0.201 & 0.153 & 0.160 & 0.123 & 0.101 & 0.099 \\ 
-%        \hline
-%        Trey Cunningham & 0.186 & 0.185 & 0.182 & 0.115 & 0.120 & 0.109 \\
-%        \hline
-%        Grant Holloway & 0.192 & 0.190 & N/A & 0.147 & 0.128 & 0.124 \\
-%        \hline
-%        Daniel Roberts & 0.181 & 0.200 & 0.183 & 0.179 & N/A & N/A \\ [0.5ex]
-%        \hline
-%       \end{tabular}
-%       \end{center}
-
-% \end{figure}
-
-Additionally it is worth noting that in the semi-finals of the World Track and
-Field Championships; Allen's reaction time was 0.101 which is only 0.001 above
-the legal limit.  What this suggests is that Allen may have strong reflexes
-and be able to react extremely well to the sound of the gun.  Devon Allen is a
-very good athlete, as evidenced by his football career at the University of Oregon
-and making it onto the Philadelphia Eagles practice squad this past year \citep{Hurley}.
-Thus the 0.099 reaction time may not have been a product of Devon Allen
-predicting the start of the race but rather a combination of a quick reaction 
-and a possibly faulty sensor.
-
+\section{Background}
+\label{sec:Background}
 Since 1985 Seiko Holding Corporations has served every World Athletics Championship
@@
 to 2022, Seiko again upgraded its technology but for its jump management system.
-\citep{Pilianidis}
+
+Something that is extremely important to note in the Data is the inconsistency of World Athletics in terms of how they penalize false starts.
+From 2007-2009, the IAAF (the former name for World Athletics, and the governing body for the World Track and Field Champions) instituted a rule change that allowed one warning false start before a sprinter was disqualified \citep{False-Start}  This essentially gave sprinters a second chance and reduced
+the penalty for a false start.  There were 18 male and 7 female false starts at the 2007
+World Championships, 18 male and 7 female starts at the 2009 World Championships.
+Starting in 2011, the rule was scrapped, and the old policy which automatically disqualifies runners who false start was reinsituted.  The effect? At the 2011 World
+Championship there were 6 male disqualifications and 4 female disqualifications \citep{False-Start} .By returning to the harsher policy and cracking down on false starts, World Athletics had reduced the number of false starts by two thirds.
+This was desirable for
+World Athletics as false starts can make already long track meets more tedious, both
+for athletes and viewers watching the television broadcast.\citep{Pilianidis}
 
 \section{Data}
-\label{sec:data}
+\label{sec:Data}
 Data was copied from the World Athletics website and pasted into an excel
@@
 was added because of suspected issues that reaction time may be correlated with other runners reacting quickly (one runner reacting quickly may result in other runners reacting quickly).
+The data that looked at in the paper is best visualized using a scatterplot and box plot.
+Figure ~\ref{fig:DataBoxPlot}
+\begin{figure}[h]
+  \centering 
+  \includegraphics[scale = 0.8]{DataBoxPlot}
+  \label{fig:DataBoxPlot}
+\end{figure}
+
+\begin{figure}[h]
+  \centering 
+  \includegraphics[scale = 0.6]{DataScatterPlot}
+  \label{fig:DataScatterPlot}
+\end{figure}
 
@@
 faster than at the USA Track Championships.  From June 23-26 2022, USA held it's Track
-and Field Championship to decide who to send to the World Championships.  Thus, a baseline
-is able to be established for the four USA athletes who competed in at least one round
-of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen. Every
+and Field Championship to decide who to send to the World Championships at Hayward Field, the exact same venue as used for the World Track and Field Championship a month later.
+Thus, a baseline is able to be established for the four USA athletes who competed in at least one round of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen.
+Every
 athlete reacted faster in all of the World Track and Field Championship races compared
@@
 \label{sec:methods}
-The main purpose of this paper is to determine whether or not Devon Allen was
-unfairly disqualified because of his reaction time.
 
 The main statistical model used in the paper is a gamma mixed effects model, which is
-in the family of Generalized Linear Mixed Model (GLMM).  A GLMM is a type of model that
-supports inputs that are from different distributions.  The model that we use is for the
+in the family of Generalized Linear Mixed Model (GLMM). The GLMM comes from the R package 
+lme4, which hosts a variety of GLMM functions \citep{Rpkg:lme4}.  A GLMM is a type of model
+that supports inputs that are from different distributions.  The model that we use is for the
 Gamma distribution for the fixed effects part of the model and a random effects part.  In
@@
 either a year effect, batch effect, or both are used.  Initially a linear mixed effects
-model was attempted to be implemented, but due to a right hand skew in the reaction times, a gamma distribution seemed better equipped to model the data (The right
-skew is not particularly suprising as it is not possible to register a reaction time below 0.0 seconds, resulting in a lower bound but no upper bound for outliers).
-In the code, the link parameter under Gamma() was changed to be "log" as logarithmic functions are better equipped to deal with high outliers.
+model was attempted to be implemented, but due to a right hand skew in the reaction times, 
+a gamma distribution seemed better equipped to model the data (The rightskew is not 
+particularly suprising as it is not possible to register a reaction time below 0.0 
+seconds, resulting in a lower bound but no upper bound for outliers). In the code, the 
+link parameter under Gamma() was changed to be "log" as logarithmic functions are better 
+equipped to deal with high outliers.  The graph below shows the skewed outliers: ~\ref{fig:Skew}.
+\begin{figure}[h]
+  \centering 
+  \includegraphics[scale = 0.6]{Skew}
+  \label{fig:Skew}
+\end{figure}
+
+Once the Gamma mixed effects model was chosen, it needed to be decided which combination
+of effects to use.  It was possible to include a year effect, a batch effect, or a year
+and batch effect.  Density plots were created that simulated the probability of a mean
+density less than 0.1, which is the current threshold for a reaction time
+disqualification.  The density plots are shown below.
+
+\begin{figure}[h]
+  \centering
+  \begin{minipage}{0.45\textwidth}
+      \centering
+      \includegraphics[width=0.9\textwidth]{Men_Density_Year} % first figure itself
+      \caption{Density plot using pooled men's data and the year effect}
+  \end{minipage}\hfill
+  \begin{minipage}{0.45\textwidth}
+      \centering
+      \includegraphics[width=0.9\textwidth]{Women_Density_Year} % second figure itself
+      \caption{Density plot using pooled women's data and the year effect}
+  \end{minipage}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \begin{minipage}{0.45\textwidth}
+      \centering
+      \includegraphics[width=0.9\textwidth]{Men_Density_Batch} % first figure itself
+      \caption{Density plot using pooled men's data and the batch effect}
+  \end{minipage}\hfill
+  \begin{minipage}{0.45\textwidth}
+      \centering
+      \includegraphics[width=0.9\textwidth]{Men_Density_Both} % second figure itself
+      \caption{Density plot using pooled men's data and both year and batch effect}
+  \end{minipage}
+\end{figure}
 
@@
 than the men's race, the reaction time data is still important and can be used to
-essentially double the data size.  However, one of the largest issues with this is that on average, men are able to react quicker than women.  Thus the data may end up following
-a bimodal distribution which would be difficult to model and not helpful for determining
-if Devon Allen's reaction time was the result of a faulty system or if it is an argument
-for chaning the 0.10 second minimum barrier that currently exists.  Thus it was decided
+essentially double the data size.
+However, one of the largest issues with this is that on average, men are able to react quicker than women.
+A study from the 2008 Beijing Olympics found that the p value of comparing the mean fastest reaction time between men and women was below 0.001 and the differnce is in part because of the greater force threshold men are able to exert on the starting blocks \citep{Lipps}.
+The data from including both men and women could follow a bimodal distribution which would be difficult to model and not helpful for determining if Devon Allen's reaction time was the result of a faulty system or if it is an argument
+for changing the 0.10 second minimum barrier that currently exists.  Thus it was decided
 to look at men's and women's data separately.  Another decision that was made was to look
@@
 be a little more anxious, more "jumpy" and that under stress and they would react faster.
-While the preliminary rounds also provide a bulk of the total observations, they are not
-as relevant to Allen's disqualification, which occured during the finals.
+Additionally, exceptionally strong atheletes may not race as hard in the early heats in
+order to reduce the risk for injury and to save themselves for the later stages of
+competition that matter more.
+While the preliminary rounds also provide a bulk of the total observations, they are not as relevant to Allen's disqualification, which occured during the finals.
+Thus, the decision was made to look at data from the men's semifinals
+and finals in a "pooled" data set.
 %%%Include picture
@@
 \label{sec:conclusion}
-%%%Need to discuss how the false start changed from 2007-2009
-The Tukey HSD performed on the finals data showed that across the many t tests it
-performed, 8 were signifcantly different of which 6 came from 2022 with the other
-two came from 2013 which was a high outlier on reaction time.  No other year had
-as low as a reaction time as 2022 for the finals, and the individual t test
-comparing 2019 and 2022 produced a p value of $2.654\cdot10^{-8}$.  This is an
-extermely significant p value and shows that there was a significant difference
-betweent the 2019 times and 2022 times, but there does not appear to be a reason
-for why the times should be so much higher.  Shoe technology may have improved
-during that time, but that would likely be reflected in the total time and not
-in the reaction time.  Additionally, other environmental factors such as wind,
-heat, etc. also should not impact how quick athletes are to react to the sound
-of the gun.  When the analysis of variance was run to compare the linear mixed
-effects model and the random effects model, the resulting p value was $3.243\cdot10^{-5}$.
-This is a highly significant p value at all alpha levels and it means that there
-does appear to be a year effect when predicting reaction time.  From the R output,
-the estimate of the intercept was 0.148883, the variance of the intercept was
-$8.643\cdot10^{-5}$, the variance of the residuals was $3.719\cdot10^{-4}$.
-However, this does not provide sufficient evidence of how strong the year effect
-is.  In order to calculate that, the pnorm() function was used, which returned
-a p value of 0.0112, which is low but not significant at all alpha levels.  This
-means that there is a very great chance that the year effect is significant but
-that we expect that if we repeatly sampled, the chance of observing a test
-statistic higher would be only 0.0112.
+It is worth noting that in the semi-finals of the World Track and
+Field Championships; Allen's reaction time was 0.101 which is only 0.001 above
+the legal limit.  What this suggests is that Allen may have strong reflexes
+and be able to react extremely well to the sound of the gun.  It seems unlikely that
+Allen would be able to correctly predict the gun with such precision two times in a row,
+which is the exact reason that the 0.10 second reaction time disqualification barrier was
+imposed in the first place.  But Allen showed that at the Championship he was able to
+react extremely fast two times, which may suggest skill rather than luck. Devon Allen is a very good athlete, as evidenced by his football career at the University of Oregon
+and making it onto the Philadelphia Eagles practice squad this past year \citep{Hurley}.
+Thus the 0.099 reaction time may not have been a product of Devon Allen
+predicting the start of the race but rather a combination of a quick reaction 
+and a possibly faulty sensor.
+
+While this sample is too small in order
+to perform a meaningful matched pairs t test, or other similar method, it is apparent
+that Allen, Cunningham, and Holloway's times are faster at the World Track and Field
+Championship compared to the United States.  There does not appear to be a particular
+reason for this, as at the World Track and Field Championship, no two United States
+athetes were in the same batch in any race.
+
+An anova test that compared the mixed effects model for only a batch effect with the
+model that included both a batch and year effect produced a p value of 0.01059, meaning
+that the model with both mixed effects was stronger, but not significantly so.  However,
+when looking at the density plots, the probability of observing a reaction time of below
+0.1 was found to be approximately zero when looking at the model that inclued both effects,
+but the probability was 0.0066 when only the batch effect was included.  Based on the density
+plot and how close the both effects model was to being statistically signficant at an alpha
+value of 0.05, the model with both the batch and year effect was chosen to be analyzed further.
+The random effects: Batch and Year both had relatively small variances: 0.003455 and 0.001714
+respectively.  For the fixed effect the estimate was -1.87968 and the p value was less than
+$2\cdot10{-16}$, which is extremely small.  The model can reparameterize the gamma distribution,
+which is calculated by $e^{-1.87968}$ which is equal to 0.152.  The variance of the gamma
+distribution is 0.0329 from the R output.  Thus a gamma distribution can be created from
+this, and the probability of a reaction time below 0.1 is essentially zero.
+
+The model accuracy improves when certain batches are removed from the data.  Batch 41 in
+particular has extremely high reaction times.  This outlier can likely be explained by a
+few different causes.
+Batch 41 is a semifinals heat of the 2013 World Championship.
+2013 had the highest average reaction time from the pooled data set.
+2013 was also the second world championship after the rule change that imposed the one strike policy on
+disqualifications. 2011 and 2013 had the highest average reaction time in the pooled
+data and were the two years after the rule change.
+What this suggests is that there could have been a psychological effect of the stricter policy that resulted in athletes being more cautious and reacting slower than they had in the previous two years under the more lax guidelines from 2007 to 2009 in order to avoid a false start and thus a disqualification.
+In Batch 41 particularly two runners did not finish, and two runners reacted extremely slow: 0.327 and 0.397 (Remember that Devon Allen's reaction time was
+0.099 seconds in the 2022 Championship final and 0.101 in the 2022 semifinal).  Removing this batch and others could greatly improve the strength of the model.  However, while
+it is a high outlier, there is not sufficient information available as to why the
+reaction times are so high, and no basis for why it should be improved other than because
+it is an outlier.
+
+\begin{figure}[h]
+  \centering 
+  \includegraphics[scale = 1]{TBY}
+  \label{fig:TBY}
+\end{figure}
+
+The below figure plots the average reaction time by year for the pooled men's data since
+1999.  The circle in the bottom right is for 2022, which is signifcantly lower than the
+data points for the previous three years, and is substially lower than the data points
+for 2011 and 2013.  The large variation is confusing because there is little explanation
+as to why there is so much of a drop off in reaction time from 2013 to 2015 or the spike
+from 2001 to 2003.
 
@@
 to conclude that Devon Allen was wrongly disqualified, that may not neccessarily
-be the case.
+be the case.  If there are issues with reaction time such as athletes approaching the 0.10 
+second barrier again at the next World Track Championship (August 2023), then World Athletics 
+needs to adjust that threshold and allow for faster reaction times so that athletes with superb 
+reflexes are not penalized like how Allen was in 2022.
 

### 2022-12-10 Code/Plots.Rmd
R code

---
title: "Plots"
author: "Owen Fiore"
date: "2022-12-09"
output: html_document
---

```{r}
times <- read.csv("../Data/PaperData.csv")
View(times)
colnames(times) <- c("Year", "Type", "Time", "RxnTime", "Gender", "Batch")
```

#DataSets
```{r}
men_finals <- subset(times, Type == "F" & Gender == "M" & Time != 'DQ') # & Year < 2022)
women_finals <- subset(times, Type == "F" & Gender == "F" & Time != 'DQ')

men_pooled <- subset(times, (Type == "F" | Type == "S") & Gender == "M" & Time != 'DQ')
women_pooled <- subset(times, (Type == "F" | Type == "S") & Gender == "F" & Time != 'DQ')
```

#Models
```{r}
library(lme4)
finals.lme <- lmer(RxnTime ~ (1|Year), data = men_pooled)
```

###Functions
```{r}
simfit <- function(fit2, n = 12000000) {
  vc <- as.data.frame(VarCorr(fit2))
  ## mu <- exp(coef(fit2)$Year[,]) ## using fitted random effects only
  z <- rnorm(n, sd = vc$sdcor[1]) #$ using generated random effects
  mu <- exp(fixef(fit2) + z)
  shape <- 1 / vc$vcov[2]
  x <- rgamma(n, shape = shape, scale = mu / shape)
  x
}

x <- simfit(fit2)
mean(x < 0.1)
quantile(x, prob = c(.00001, .0001, .0005, .001))
## compare with and without random effect
plot(density(x)) # with random effect
curve(dgamma(x, shape = 1 / sfit0$dispersion,
             scale = exp(coef(fit0)) * sfit0$dispersion),
      0.08, 0.22, add = TRUE, col = "blue") # fixed effect only
```

#Figures
```{r}
pdf(file = "TBY.pdf")
plot(aggregate(men_pooled$RxnTime, list(men_pooled$Year), FUN = mean), xlab = "Year", ylab = "Reaction time in seconds", main = "Mean Reaction Time By Year")
dev.off()
```

```{r}
pdf(file = "Skew.pdf")
hist(resid(finals.lme), xlab = "Residuals of Linear mixed effects model", main = "Right skew when plotting the residuals of the linear mixed effects model")
dev.off()
```
```{r}
pdf(file = "Gamma_skew.pdf")
dev.off()
```

```{r}
pdf(file = "DataBoxplot.pdf")
boxplot(RxnTime ~ Year, data = men_pooled, xlab = "Year", ylab = "Reaction Time in seconds", main = "Men's Semifinal and Final Reaction Time Boxplot Data")
dev.off()
```

```{r}
pdf(file = "DataScatterplot.pdf")
plot(RxnTime ~ Year, data = men_pooled, xlab = "Year", ylab = "Reaction Time in seconds", main = "Men's Semifinal and Final Reaction Time Scatterplot Data")
dev.off()
```

```{r}
set.seed(1)
library(lme4)
pdf(file = "Men_Density_Year.pdf")
fit2 <- glmer(RxnTime ~ (1 | Year), data = men_pooled, family = Gamma(link = "log"))
fit0 <- glm(RxnTime ~ 1, data = men_pooled, family = Gamma(link = "log"))
sfit0 <- summary(fit0)
simfit <- function(fit2, n = 12000000) {
  vc <- as.data.frame(VarCorr(fit2))
  ## mu <- exp(coef(fit2)$Year[,]) ## using fitted random effects only
  z <- rnorm(n, sd = vc$sdcor[1]) #$ using generated random effects
  mu <- exp(fixef(fit2) + z)
  shape <- 1 / vc$vcov[2]
  x <- rgamma(n, shape = shape, scale = mu / shape)
  x
}

x <- simfit(fit2)
mean(x < 0.1)
quantile(x, prob = c(.00001, .0001, .0005, .001))
## compare with and without random effect
plot(density(x), main = "Men's Pooled With Only Year Effect") # with random effect
curve(dgamma(x, shape = 1 / sfit0$dispersion,
             scale = exp(coef(fit0)) * sfit0$dispersion),
      0.08, 0.22, add = TRUE, col = "blue") # fixed effect only
legend("right", title = "Probability mean less than", legend = " 0.1 is 0.006847")
dev.off()
```

```{r}
set.seed(1)
library(lme4)
pdf(file = "Men_Density_Batch.pdf")
fit2 <- glmer(RxnTime ~ (1 | Batch), data = men_pooled, family = Gamma(link = "log"))
fit0 <- glm(RxnTime ~ 1, data = men_pooled, family = Gamma(link = "log"))
sfit0 <- summary(fit0)
simfit <- function(fit2, n = 12000000) {
  vc <- as.data.frame(VarCorr(fit2))
  ## mu <- exp(coef(fit2)$Year[,]) ## using fitted random effects only
  z <- rnorm(n, sd = vc$sdcor[1]) #$ using generated random effects
  mu <- exp(fixef(fit2) + z)
  shape <- 1 / vc$vcov[2]
  x <- rgamma(n, shape = shape, scale = mu / shape)
  x
}

x <- simfit(fit2)
mean(x < 0.1)
quantile(x, prob = c(.00001, .0001, .0005, .001))
## compare with and without random effect
plot(density(x), main = "Men's Pooled With Only Batch Effect") # with random effect
curve(dgamma(x, shape = 1 / sfit0$dispersion,
             scale = exp(coef(fit0)) * sfit0$dispersion),
      0.08, 0.22, add = TRUE, col = "red") # fixed effect only
legend("right", title = "Probability mean less than", legend = " 0.1 is 0.0066115")
dev.off()
```

```{r}
set.seed(1)
library(lme4)
pdf(file = "Men_Density_Both.pdf")
fit2 <- glmer(RxnTime ~ (1 | Year) + (1 | Batch), data = men_pooled, family = Gamma(link = "log"))
fit0 <- glm(RxnTime ~ 1, data = men_pooled, family = Gamma(link = "log"))
sfit0 <- summary(fit0)
simfit <- function(fit2, n = 12000000) {
  vc <- as.data.frame(VarCorr(fit2))
  ## mu <- exp(coef(fit2)$Year[,]) ## using fitted random effects only
  z <- rnorm(n, sd = vc$sdcor[1]) #$ using generated random effects
  mu <- exp(fixef(fit2) + z)
  shape <- 1 / vc$vcov[2]
  x <- rgamma(n, shape = shape, scale = mu / shape)
  x
}

x <- simfit(fit2)
mean(x < 0.1)
quantile(x, prob = c(.00001, .0001, .0005, .001))
## compare with and without random effect
plot(density(x), main = "Men's Pooled With Year and Batch Effect") # with random effect
curve(dgamma(x, shape = 1 / sfit0$dispersion,
             scale = exp(coef(fit0)) * sfit0$dispersion),
      0.08, 0.22, add = TRUE, col = "orange") # fixed effect only
legend("right", title = "Probability mean less than", legend = " 0.1 is 0")
dev.off()
```
```{r}
set.seed(1)
library(lme4)
pdf(file = "Women_Density_Year.pdf")
fit2 <- glmer(RxnTime ~ (1 | Year), data = women_pooled, family = Gamma(link = "log"))
fit0 <- glm(RxnTime ~ 1, data = women_pooled, family = Gamma(link = "log"))
sfit0 <- summary(fit0)
simfit <- function(fit2, n = 12000000) {
  vc <- as.data.frame(VarCorr(fit2))
  ## mu <- exp(coef(fit2)$Year[,]) ## using fitted random effects only
  z <- rnorm(n, sd = vc$sdcor[1]) #$ using generated random effects
  mu <- exp(fixef(fit2) + z)
  shape <- 1 / vc$vcov[2]
  x <- rgamma(n, shape = shape, scale = mu / shape)
  x
}

x <- simfit(fit2)
mean(x < 0.1)
quantile(x, prob = c(.00001, .0001, .0005, .001))
## compare with and without random effect
plot(density(x), main = "Women's Pooled With Only Year Effect") # with random effect
curve(dgamma(x, shape = 1 / sfit0$dispersion,
             scale = exp(coef(fit0)) * sfit0$dispersion),
      0.08, 0.22, add = TRUE, col = "deeppink") # fixed effect only
legend("right", title = "Probability mean less than", legend = " 0.1 is 0.0042296")
dev.off()
```
```{r}
year_effect <- glmer(RxnTime ~ (1 | Year), data = men_pooled, family = Gamma(link = "log"))
batch_effect <- glmer(RxnTime ~ (1 | Batch), data = men_pooled, family = Gamma(link = "log"))
both <- glmer(RxnTime ~ (1 | Year) + (1 | Batch), data = men_pooled, family = Gamma(link = "log"))
#anova(batch_effect, both)
summary(both)
```

```{r}
t.test(men_pooled$RxnTime[men_pooled$Year==2019],men_pooled$RxnTime[men_pooled$Year==2022])
```
